Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems

### 2023-12-08 2024pipit/Main.tex
start pipit paper

%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
\documentclass[sigplan,screen, review]{acmart}

% any of these necessary?
% \usepackage{amssymb}
\usepackage{amsthm}
% \usepackage{graphicx}
\usepackage{amsmath}
% \usepackage{mathptmx}
\usepackage{mathtools}
% \usepackage{stmaryrd}
% \usepackage{adjustbox}
% \usepackage{hyperref}
% \usepackage{alltt}
% \usepackage{url}
% \usepackage{float}
% \usepackage{minipage-marginpar}

@@

\setcopyright{acmcopyright}
\copyrightyear{2023}
\acmYear{2023}
\acmDOI{XXXXXXX.XXXXXXX}

\acmConference[TyDe '23]{In submission}{2023}{In submission}
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
% \acmPrice{15.00}
% \acmISBN{978-1-4503-XXXX-X/18/06}
%%\acmSubmissionID{123-A56-BU3}

@@

\title{Pipit: Reactive Systems in \fstar ~(Extended Abstract)}

\newcommand{\affiliationANU}{
  \affiliation{%
  \institution{Australian National University}
  \city{Canberra}
  \country{Australia}
}
}

\author{Amos Robinson}
% \authornote{}
\email{amos.robinson@anu.edu.au}
\orcid{0009-0004-4837-4981}
\affiliationANU{}
\author{Alex Potanin}
% \authornotemark[1]
\email{alex.potanin@anu.edu.au}
\orcid{0000-0002-4242-2725}
\affiliationANU{}

% \renewcommand{\shortauthors}{Robinson et al.}

@@

%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
\begin{CCSXML}
<ccs2012>
    <concept>
        <concept_id>10010520.10010553.10010562.10010564</concept_id>
        <concept_desc>Computer systems organization~Embedded software</concept_desc>
        <concept_significance>300</concept_significance>
        </concept>
    <concept>
        <concept_id>10010520.10010570.10010572</concept_id>
        <concept_desc>Computer systems organization~Real-time languages</concept_desc>
        <concept_significance>500</concept_significance>
        </concept>
    <concept>
        <concept_id>10003752.10010124.10010138.10010142</concept_id>
        <concept_desc>Theory of computation~Program verification</concept_desc>
        <concept_significance>500</concept_significance>
        </concept>
    <concept>
        <concept_id>10003752.10003790.10003793</concept_id>
        <concept_desc>Theory of computation~Modal and temporal logics</concept_desc>
        <concept_significance>100</concept_significance>
        </concept>
    <concept>
        <concept_id>10011007.10011006.10011050.10011023</concept_id>
        <concept_desc>Software and its engineering~Specialized application languages</concept_desc>
        <concept_significance>500</concept_significance>
        </concept>
  </ccs2012>
\end{CCSXML}

\ccsdesc[300]{Computer systems organization~Embedded software}
\ccsdesc[500]{Computer systems organization~Real-time languages}
\ccsdesc[500]{Theory of computation~Program verification}
\ccsdesc[100]{Theory of computation~Modal and temporal logics}
\ccsdesc[500]{Software and its engineering~Specialized application languages}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Lustre, streaming, reactive, control}

\received{1 June 2023}
% \received[revised]{12 March 2009}
% \received[accepted]{5 June 2009}

@@

\section{Introduction}
Safety-critical control systems, such as the anti-lock braking systems that are present in most cars today, need to be correct and execute in real-time.
One approach, favoured by parts of the aerospace industry, is to implement the controllers in a high-level language such as Lustre~\cite{caspi1995functional} or Scade~\cite{colaco2017scade}, and verify that the implementations satisfy the high-level specification using a model-checker, such as Kind2~\cite{champion2016kind2}.
These model-checkers can prove many interesting properties automatically, but do not provide many options for manual proofs when the automated proof techniques fail.
Additionally, the semantics used by the model-checker may not match the semantics of the compiled code, in which case properties proved do not necessarily hold on the real system.
This mismatch may occur even when the compiler has been verified to be correct, as in the case of Vélus~\cite{bourke2017formally}.
For example, in Vélus, integer division rounds towards zero, matching the semantics of C; however, integer division in Kind2 rounds to negative infinity, matching SMT-lib~\cite{BarFT2016SMTLIB,kind2023intdiv}.

To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
In this abstract we introduce Pipit\footnote{Implementation available at \url{https://github.com/songlarknet/pipit}}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
Pipit aims to provide a high-level language based on Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
% By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
Pipit translates its expression language to a transition system for k-inductive proofs, which is verified; verifying the translation to imperative code is ongoing.

In this extended abstract we briefly describe the following preliminary results, which we intend to describe fully in a future publication:

\begin{itemize}
  \item we motivate Pipit which, as an embedded language, provides syntactic convenience with a small verifiable core language (\autoref{s:tut});
  \item we demonstrate the use of \fstar{}'s existing normalisation and proof automation to prove certain properties with minimal effort (\autoref{ss:tut:bibo});
  \item we describe a key difference between Pipit's core language and Lustre (\autoref{s:core}); and
  % \item we demonstrate the use of normalisation, tactics and \lowstar{} extraction to generate unverified executable code (\autoref{s:extraction}); and
  \item we evaluate Pipit by executing a verified controller on an embedded system (\autoref{s:coffee-machine}).
\end{itemize}

%CUT
% \begin{itemize}
%   \item We identify the invariants and lemmas required to prove correctness of the translation from a clock-free Lustre-style language to a transition system (\autoref{ss:meta-theory}).
%   \item We present a syntax-directed operational semantics for a clock-free Lustre-style language (\autoref{ss:operational-semantics}).
%     % TODO: CUT WAFFLE
%     This syntax-directed semantics is in contrast to previous \emph{relational} operational semantics of Lustre, which are not syntax-directed and require detailed proofs of determinism.
%     Proving determinism on our semantics requires only straightforward induction; we believe that this semantics may be better for proving other properties too.
%   \item We show preliminary results that demonstrate the effectiveness of Pipit by implementing a verified controller for a simple embedded system (\autoref{s:coffee-machine}).
% \end{itemize}

% \begin{itemize}
%   \item an embedded language makes it easier to express interesting programs with a small real-time synchronous core language;
%   \item by embedding in a language with good proof automation support, we can reason about embedded programs without much manual proof effort;
%   \item by embedding in a language with manual proof support, we can prove harder problems than existing model checkers; and
%   \item by embedding in a language with a well-defined extraction to (heap) allocation-free C code, we can get a partially verified compiler; so far we have only proved type preservation
% \end{itemize}

@@

\input{figures/core-grammar.tex}
% \input{figures/core-typing.tex}
% \input{figures/core-bigstep.tex}

@@

Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
% To ensure that streams are productive,
As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \citet{caspi1995functional} which is not syntax-directed.
Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax at the expense of potentially duplicating expressions.

@@

% \pagebreak

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{Main}

@@

\end{document}
\endinput
%%
%% End of file `sample-sigplan.tex'.

>>>>

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
% To anonymise:
% \documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate,anonymous]{lipics-v2021}

% \pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

\bibliographystyle{plainurl}

% is this package necessary?
\usepackage{mathtools}

@@

\title{Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems}

\author{Amos Robinson}{Australian National University, Canberra, Australia}{amos.robinson@anu.edu.au}{https://orcid.org/0009-0004-4837-4981}{}

\author{Alex Potanin}{Australian National University, Canberra, Australia}{alex.potanin@anu.edu.au}{https://orcid.org/0000-0002-4242-2725}{}

\authorrunning{A. Robinson and A. Potanin}
\Copyright{Amos Robinson and Alex Potanin}
% \ccsdesc[300]{Computer systems organization~Embedded software}
\ccsdesc[500]{Computer systems organization~Real-time languages}
\ccsdesc[500]{Theory of computation~Program verification}
% \ccsdesc[100]{Theory of computation~Modal and temporal logics}
\ccsdesc[500]{Software and its engineering~Specialized application languages}
\keywords{Lustre, streaming, reactive, verification}

% \category{}

@@

@@

@@

\input{section/01-Introduction.tex}

@@

\input{figures/core-grammar.tex}
\input{figures/core-typing.tex}
\input{figures/core-bigstep.tex}

@@

Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
% To ensure that streams are productive,
As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \cite{caspi1995functional} which is not syntax-directed.
Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax at the expense of potentially duplicating expressions.

@@

\bibliography{Main}

@@

\end{document}
\endinput
